---
title: "STAT216-Final-Project"
author: "Nicole Cassara"
date: "2023-12-05"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, tidy = TRUE)
```

Load in the dataset and store as object shelter:
```{r}
shelter <- read.csv("HomelessShelter.csv")
summary(shelter)
```

**Case QC Analysis: Age and Gender Analysis**

Age: Quantitative Variable
Gender: 
```{r}
#T-test to compare mean age between different genders
t.test(AGE ~ GENDER, data = shelter)

#ANOVA to test for significant age differences across different gender categories
anova(aov(AGE ~ GENDER, data = shelter))

```


$H_0$: There is no significant difference in the mean age between different genders among shelter-seekers.  
$H_1$: There is a significant difference in the mean age between at least two gender categories among shelter-seekers.

$H_0: \mu_M - \mu_F = 0$  
$H_1: \mu_M - \mu_F \ne 0$  


*ANOVA Hypotheses:*

$H_0$: There is no significant difference in the mean age across different gender categories among shelter-seekers.

$H_1$: There is a significant difference in the mean age across at least two gender categories among shelter-seekers.

Normality Assumption visualization using ggplot:
```{r}
library(ggplot2)

#Histogram and QQ plot for age within each gender category
ggplot(shelter, aes(x = AGE, fill = GENDER)) +
  geom_histogram(position = "identity", bins = 30, alpha = 0.7) +
  facet_wrap(~GENDER, scales = "free") +
  ggtitle("Distribution of Age by Gender") +
  xlab("Age") +
  ylab("Frequency")

#QQ plot
ggplot(shelter, aes(sample = AGE, color = GENDER)) +
  stat_qq() +
  stat_qq_line() +
  ggtitle("Q-Q Plot of Age by Gender") +
  xlab("Theoretical Quantiles") +
  ylab("Sample Quantiles")

#Formal test for normality (Shapiro-Wilk test for each gender category)
shapiro_test_result <- by(shelter$AGE, shelter$GENDER, shapiro.test)
print(shapiro_test_result)
```

Using base R
```{r}
# Histogram for age within each gender category
par(mfrow=c(1, 2))
hist(shelter$AGE[shelter$GENDER == "Male"], main="Histogram - Male", xlab="Age", col="lightgreen")
hist(shelter$AGE[shelter$GENDER == "Female"], main="Histogram - Female", xlab="Age", col="purple")

#QQ plot
par(mfrow=c(1, 1))
qqnorm(shelter$AGE)
qqline(shelter$AGE)

table(shelter$GENDER)

shapiro.test(shelter$AGE[shelter$GENDER == "Male"])
shapiro.test(shelter$AGE[shelter$GENDER == "Female"])

#round(stat.desc(x=shelter$AGE, norm=TRUE), 3)
```

Homogeniety of Variance
```{r}
library(car)
leveneTest(shelter$AGE, shelter$GENDER)
```

Wilcox Rank Sum Test:
```{r}
wilcox.test(AGE ~ GENDER, data = shelter)
```

Case CC-Substance Abuse and Probation Status
Chi-Sq Test of Independence

Binary: 1 is true, 0 is false

$H_0$: Probation status and substance abuse are independent

$H_1$: Probation status and substance abuse are associated

$H_0: p_1 - p_2 = 0$  
$H_1: p_1 - p_2 \ne 0$  

```{r}
tab1 <- table(shelter$probation, shelter$substanceabuse)
tab1

chisq.test(tab1, correct = F)
```
**With a p-value of 0.03481, which is below our alpha of 0.05, we can reject our null. We have sufficient evidence to conclude that substance abuse and probation status are associated.**


Case QQ: Regression
```{r}
lm1 <- lm(NIGHTS ~ AGE, data = shelter)
summary(lm1)

plot(shelter$NIGHTS ~ shelter$AGE)
abline(lm1) #plots the regression line

##the line is y = 264.2178 - 3.9619x

##The intercept, 264.2178, is the expected value of y (nights) when x (age) equals 0.

##The slope, -3.9619, is the expected change in y (nights) for 1 unit increase in x (age).

##Coefficient of Determination: 0.4053
```


```{r}
shapiro.test(shelter$AGE)
shapiro.test(shelter$INCOME)

hist(shelter$AGE)
hist(shelter$INCOME)

qqnorm(shelter$AGE, pch = 1, frame = FALSE)
qqline(shelter$AGE, col = "steelblue", lwd = 2)

qqnorm(shelter$INCOME, pch = 1, frame = FALSE)
qqline(shelter$INCOME, col = "steelblue", lwd = 2)
```
Case QQ+
```{r}
shelter$GENDER <- as.factor(shelter$GENDER)
shelter$VETERAN <- as.factor(shelter$VETERAN)

lm1 <- lm(INCOME ~ AGE + GENDER + VETERAN + NIGHTS + substanceabuse + completed + probation + assistance_type, data = shelter)
summary(lm1)
```
